\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-10-04 writing/manuscript.Rmd
mass push to transfer to laptop

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emilia Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
    # - \usepackage{float} # for better floating objects interface; adds more floats 
    # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    keep_tex: true
    number_sections: false
    toc: false
---

<!--
TO DO:
- write conclusion paragraph
- write discussion (2-3 pages)
    - add headers for sections
- trim down and edit intro and other previous sections
- move empirical examples within each relevant section (i.e., abundance examples for abundance section)
- simple figure for demonstrating idea of nonlinear relationship between R and H, with animal moving in empty space?
    - explain the idea in the caption, straight line can't work because can't have negative space, and there's need for saturation
    - line increasing as R decreases, 3 options: death, more increase (nomadism), sub-optimal saturation
- make same figure for V(R), but focus on temporal variance in time:
    - use grey or low alpha for curves with high uncertainty
- focus on V(R) as temporal variance rather than spatial one. people have already focused on patchiness, but temporal variation is more risky
- drop figure 1? (Fictitious example of variation)
- move 5x5 figure to interactive effects section
- add median NDVI behind tracking data of tapir, buffalo, and ??? figure
-->

@@

@@

```{r wrap-figure-function, include = FALSE}
# code modified from https://stackoverflow.com/questions/54100806/r-markdown-how-do-i-make-text-float-around-figures

# this chunk is needed to wrap text around or beside figures

@@

<!--
@ciuti and @berger-tal: human presence in nature for other reasons than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
habitat productivity hypothesis: Harestad and Bunnel, 1979
-->

@@

Animal's spatial needs are often thought to be tighly linked to the amount of resources ($R$) within their habitats. Although this hypothesis has widespread support, most studies only account for the mean $R$, while the variance in $R$ is rarely accounted for. Failing to account for such variation is likely to result in an underestimation on animals' spatial needs, which may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict. In this paper, we provide a review of animal movement studies that estimate animals' spatial needs as a function of $R$, and we demonstrate the importance of accounting for variance in $R$. Additionally, we provide a unifying framework and hypotheses for the effect of mean and variance in $R$ on animals' home range sizes and movement behaviors. In the first part of this paper, we use simulated movement data in a variety of example habitats to demonstrate the importance of accounting for the variance in $R$. Next, we use the Normalized Difference Vegetation Index to estimate mean and variance in $R$ within the habitats of a tapir from the Pantanal (Brazil) and an African buffalo from South Africa to demostrate how this framework can be applied to real-world tracking data. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to real-world tracking data.

@@

The ability to move allows motile organisms to respond rapidly and continuously to heterogeneous and changing environments and conditions [@southwood_habitat_1977]. Animals may move for a variety of reasons, including searching for resources [@refs] or mates [@refs] while avoiding predation [@refs] or dangerous competition [@refs], as well as defending resources and territories [@grant_whether_1993; @refs]. @nathan_movement_2008 conceptualized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model

@@

such that the animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters, doing so with sufficient frequency and accuracy would not be simple. Rather than measuring changes in the various resources and needs directly, we can quantify animals' spatial needs as a response to average resource abundance and the variability around it, instead. This paper will illustrate how animals' spatial needs depend on average resource abundance and the variance around it. We present this idea through a quantitative theoretical framework and empirical examples where resource abundance is estimated using the Normalized Vegetation Index [NDVI, see @pettorelli_normalized_2011].

@@

Figure 2 of @bista_effects_2022 for example of study that assumes linear relationships (w LM) for home range, when there can't be one.

```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of mean resource abundance ($\\mathbb E(R)$). At low values of $\\mathbb E(R)$, range-resident animals (bottom dashed line) may exhibit nomadic behavior, such that While $H$ decreases as $\\mathbb E(R)$ increases, it is not possible for $H$ to continuously decrease linearly because animals will require a minimum (nonnegative) amount of space to survive. At low values of $\\mathbb E(R)$, $H$", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

Environmental productivity is tightly linked to the amount of space animals need to cover to obtain the resources they needed to survive and reproduce [figure \@ref(fig:), @relyea_home_2000]. While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction, such as energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @refs], and maintaining a thermoregulatory balance. The size of a home range is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005]. Animals living in large groups will require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], and larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004].

The favorableness of a patch or habitat often depends on a variety of factors, including resource abundance, competitive pressure, and predation risk. Since many of these factors' affect resource abundance unpredictably, it is useful to visualize it as a random variable^[In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$, $H$). Known values, such as realizations (i.e., known observations) of random variables, are indicated with lower-case letters (e.g., $r$, $h$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.]. Let $R$ indicate the amount of *resources* in a particular patch or area. For simplicity, we can assume $R$ can take any value between zero and infinity, and we can let $R$ follow a Gamma distribution with mean $\mu$ and variance $\sigma^2$ (which we can write as $R \sim \Gamma(\mu, \sigma^2)$)^[Although Gamma distributions are more often parameterized using shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ for simplicity. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean since as the mean approaches zero the variance also does: $\mu = 0 \iff k \theta = 0 \iff k \theta^2 = 0 \iff \sigma^2 = 0$.]. Next, let $H$ indicate the size of an animal's home range (HR)^[We can consider specific utilization quantiles of the HR, such as the core HR, $H_{50\%}$, or the 95% HR, $H_{95\%}$, but for simplicity I will refer to the entirety of the HR with $H$.].

>>>>

---
title: \large How resource abundance and stochasticity affect animals' spatial needs
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emilia Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
    # - \usepackage{float} # for better floating objects interface; adds more floats 
    # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
---

<!--
TO DO:
- write conclusion paragraph
- write discussion (2-3 pages)
  - add headers for sections
- focus on V(R) as temporal variance rather than spatial one. people have already focused on patchiness, but temporal variation is more risky
-->

@@

\tableofcontents

@@

```{r wrap-figure-function, include = FALSE}
# this chunk is needed to wrap text around or beside figures
# code modified from https://stackoverflow.com/questions/54100806/r-markdown-how-do-i-make-text-float-around-figures

@@

<!--
@ciuti and @berger-tal: human presence in nature for other reasons than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
habitat productivity hypothesis: Harestad and Bunnel, 1979 [@harestad_home_1979]
-->

@@

\noindent Since animals need to maintain a positive energetic balance to grow and reproduce, their spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats. Although this hypothesis has widespread empirical support, existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance in around the mean are still largely unknown. This is not a trivial oversight. Animals adjust to variable environmental conditions, so failing to account for the effect of resource unpredictability can result in a limited understanding of animals' spatial needs. In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resource abundance on animals' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use. Finally, we use real-world tracking data on a lowland tapir (\textit{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

@@

<!-- You need a lot more depth here. Open with a paragraph on movement generally that leads to the idea that responses to resource are something worth focusing on. Follow that up with a paragraph on the history of exploring movement/space use and resources, leading to the concept that, to date, the relationship between abundance and stochasticity have not been fully explored. Then finish with a paragraph detailing the aims of the current paper. -->

\noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019], or even death [@foley_severe_2008; @berger_climate_2018]. Thus, many organisms will change their behaviours in response to changes in local resource abundance. While some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020]. However, movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with larger decreases when resources are scarce [e.g., @herfindal_prey_2005; @nilsen_can_2005]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago various scientists, inlcuding @lindstedt_home_1986 and @reiss_scaling_1988. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.

In this study, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive chapters. Concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology. Next, we provide a review of the effects of resource abundance on animals spatial needs while suggesting a unifying hypothesis of how animals respond to changes in resource abundance. Afterwards, we present a hypothesis for how animals respond to resource unpredictability. Subsequently, we test the two hypotheses using quantitative simulated animal responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir living in the Brazilian Cerrado. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

# Concepts and definitions

## Resources as a random variable

In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations (i.e., known observations or instances) of random variables, are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Since resource abundance depends on a variety of factors, including climate, weather, competitive pressure, and individual- and species-specific energetics, it is often unpredictable (and hard to quantify). Thus, we can let the random variable $R$ indicate the *resource* abundance at a given point in space and time.

### Probability distributions

In probability and statistics, random variables are defined by specifying the distribution the variable follows. Since the variable is random, it can take multiple values (minimum 2), each with different probabilities. Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution. The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**. The support of the normal and T distributions are all real numbers, while the Bernoulli distribution only has support over values 0 and 1.

There are many distributions we can assign to $R$, depending on how we quantify it. If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow any distribution with support over all real numbers between 0 and infinity, which we can indicate with $r \in (0, \infty)$. (We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because both 0 and infinity are not included in the support set.) In this case, we can let $R$ follow a Gamma distribution. If $R$ was a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution for $R$. Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). The next two sections introduce two parameters that are often used to visualize and define distributions: the expectation and the variance of a random variable.

### Expected resource abundance, $\mathbb E(R)$

The expectation of a random variable (i.e. its mean or average, indicated here as ) is the value one can *expect* the random variable to take. $\mathbb E(R) = \mu$ to indicate the expectation of the random variable $R$. The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\mathbb E(R)$ is still useful for estimating the average long-term resource abundance. When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\mathbb E(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\mathbb E(R) = \mu(t)$.

The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @bista_effect_2022]. This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero). The relationship between resource abundance and animals' spatial needs is explored in more depth in a following section.

### Variance in resource abundance, $\mathbb V(R)$

A random variable's variance is a measure of its unpredictability or variability. We use  $\mathbb V(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$. For instance, while one may consider $R$ to vary within years, we define $\mathbb V(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others. Additionally, we distinguish between two different sources of variation. We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley. In contrast, whether next year will be a good or bad year for berries is stochastic.

### Simulating resource abundance

In this paper, we can simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$). Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean. As the mean approaches zero (from the positive side, which we write as $\mu \rightarrow0^+$) the variance also does: $\mu \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \sigma^2 = 0^+$.

### Estimating resource abundance using NDVI

In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011]. However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the linear transformation

\begin{equation} \label{ndvi-to-r}
R = \frac{\nu + 1}{2},
\end{equation}

where $\nu$ indicates the NDVI value. Note that only **linear transformations** (i.e. only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution's parameters [@denny_fallacy_2017; @jensen_sur_1906]. In this case, we can define $R$ as following a Beta distribution using the notation $B (\mu(t), \sigma^2(t))$. Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.

## Animal movement as a stochastic process

### Drivers of animal movement

@nathan_movement_2008 generalized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model

@@

such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec r_t, \text{and }\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels, @refs), doing so with sufficient frequency and accuracy would not be simple. However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly) <!-- since accounting for a proxy of resource abundance will also account for the effects of many other drivers that depend strongly on resource abundance (e.g., competition, predation)-->.

### Visualizing movement as a stochastic process

To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day. If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another. However, we can use the animal's positional data to estimate the amount of time it spends exploring before returning "home" (or the time required for the position to "decay" back to the average position), which is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$. Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity. $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter. Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters. Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].

If sampling frequency is too low to provide precise measures of autocorrelation, it will be inappropriate to estimate an animal's average speed, directional persistence, and range crossing time, but the data can still provide reasonable estimates of the animal's spatial needs and usage, and model selection procedures with criteria such as AIC can help the analyst choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022]. Estimates of average or instantaneous speed require the highest sampling frequency [$\Delta t \lessapprox \tau_v$, see @noonan_scale-insensitive_2019], since the dataset must contain multiple consecutive samples where the animal is moving with a similar speed and direction. However, since the time required for an animal to cross the area it uses is orders of magnitude larger than the time it moves with a consistent behavior, it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed. Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home range size, provided that the sampling duration is sufficiently long.

### Quantifying spatial needs

The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Thus, here we quantify animal's spatial needs using their positional variance (indicated as $\mathbb V(\text{position})$), instead. ***explain the concept of positional variance*** [@refs].

Although range-resident animals will generally have smaller $\mathbb V(\text{position})$ than migratory and nomadic animals, $\mathbb V(\text{position})$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\mathbb V(\text{position})$ determines the scale and size of an animal's spatial needs but not its behavior. Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different kinds of movement. Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year). In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it is unlikely to return to locations it visited in the past.

### Advantages of modeling animal movement as a stochastic process

Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement.

Depending on sampling frequency, statistical methods that rely on straight-line displacement are prone to under-estimating the distance traveled by an animal and its speed at coarse sampling frequencies, while they over-estimate them at fine frequencies [@noonan_scale-insensitive_2019]. In addition, most commonly-used methods for home range estimation assume tracking data to be independent of one another and sampled at consistent and discrete intervals, which is most often not the case. Consequently, animals' spatial needs are often under-estimated [@noonan_comprehensive_2019].

### Unassigned

@teitelbaum_beyond_2019:

- **Nomadic movement**: movement with irregular timing and/or direction that produces both within- and between-year variability in location and movement patterns
- "Data from recent developments in animal tracking technology (e.g., longer-lasting and lighter GPS devices) show that species previously considered migrants or range residents can show irregular, often long-distance, movement patterns (flexible foraging movements of leatherback turtles across the north Atlantic ocean, To migrate, stay put, or wander Varied movement strategies in bald eagles, Big data analyses reveal patterns and drivers of the movements of southern elephant seals)"

@@

```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\mathbb V(\\text{position})$) of an animal as a function of mean resource abundance ($\\mathbb E(R)$). We expect low values of $\\mathbb E(R)$ to result in high $\\mathbb V(\\text{position})$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\mathbb E(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive. Note that the relationship between $\\mathbb E(R)$ and $\\mathbb V(\\text{position})$ cannot be linear because it would require $\\mathbb V(\\text{position})$ to be negative for high values of $\\mathbb E(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

<!-- You need to open with some background on metabolic requirements. Individuals have fixed needs that they must meet in order to survive, grow, and reproduce. If these are not met, they will die. Then transition to the idea that, because of that, they move around in order to secure resources. Therefore, when environments are productive, they don't have to move as far to meet their energetic requirements. -->

* add refs to generalize beyond terrestrial mammals

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @ref]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005; @ex; see figure \@ref(fig:mu-hyp)]. Low values of $\mathbb E(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp). As $\mathbb E(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022]. It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between the types of movement is important for quantifying the effect of resource abundance on animal's spatial needs. From a quantitative perspective, the switch is related to an animal's home range crossing time (or positional autocorrelation, $\tau_p$). As the amount of space an animal uses over its lifespan, the amount of time required to cross the area ($\tau_p$) will also increase, so animals with $\tau_p$ on the order of their expected lifespan will necessarily be nomadic.

```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\mathbb V(\\text{position})$) of an animal as a function of resource unpredictability ($\\mathbb V(R)$). We expect low values of $\\mathbb V(R)$ to result in low $\\mathbb V(\\text{position})$ as animals are able to depend on somewhat predictable resources. As $\\mathbb V(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration. Note that the relationship between $\\mathbb V(R)$ and $\\mathbb V(\\text{position})$ cannot be linear because it would require $\\mathbb V(\\text{position})$ to be negative for low values of $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
```

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\mathbb V(R)$ has received far less attention [but see @lucherini_habitat_1996; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\mathbb V(R)$ can change strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018]. Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs. A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.

Simlarly to the effect of $\mathbb E(R)$, we expect $\mathbb V(R)$ to have a nonlinear effect on $\mathbb V(\text{position})$. When $\mathbb V(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\mathbb V(R)$ to have a noticeable effect on $\mathbb V(\text{position})$. However, as resources become additionally unpredictable, we expect spatial needs to increase superlinearly. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors, including nomadism, dispersal, and migration (figure \@ref(fig:s2-hyp)).
